---
title: "Rreproducible Research Course Project 1"
author: "Brendon Ushe"
date: "July 6, 2017"
output: 
  html_document: 
    keep_md: true 
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

#1. Loading and preprocessing the data
###1.1. Read the data from the GitHub repository. Dimensional exploration.

```{r ReadData}
setwd("C:/ReproducibleResearch")
rdata<-read.csv("activity.csv")
head(rdata)
sapply(rdata, class)
```


###1.2. Process/transform the data (if necessary) into a format suitable for your analysis.

Tidy the data or preprocess the data

We convert the date field to Date class and interval field to Factor class.

rdata$date is converted from factor to date class. I used the "zoo" package to manage the time serie which, furthermore, is regularly spaced. The data are collected on 61 days.


```{r format}

rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
str(rdata)
unique(rdata$date)

```


#2. What is mean total number of steps taken per day?

###2.1 Calculate the total number of steps taken per day
The result is calculated and stored in the object "steps_day". Missing values are removed. Since the resulting table is quinte large, I set {r, results = "hide"}. Approximately the 13% of cases in "steps" result in missing data.


```{r MeanCalcu}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)

```


###2.2. Make a histogram of the total number of steps taken each day

Now we make a histogram of the total number of steps taken per day, plotted with appropriate bin interval.


```{r PlotHistogram}
library(ggplot2)

ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw()

```


### 2.3 Now we calculate the mean and median of the number of steps taken per day.

```{r MeanMedian}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)

steps_mean

steps_median

```

Mean number of steps taken per day is 1.076618910^{4}.

Median number of steps taken per day is 10765.

#3. What is the average daily activity pattern?

```{r AverageDailyPatterns}

#We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.

steps_per_interval <- aggregate(rdata$steps, 
                                by = list(interval = rdata$interval),
                                FUN=mean, na.rm=TRUE)

#convert to integers this helps in plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")

steps_per_interval <- aggregate(rdata$steps, 
                                by = list(interval = rdata$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")

```


### 3.1 We make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals:

```{r AverageDailyActivityPatternPlot}

ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="orange", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()

```


### 3.2 Now, we find the 5-minute interval with the containing the maximum number of steps:

```{r 5mininterval}

max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]

```

# 4.0 Imputing missing values:

### 4.1. Total number of missing values:

```{r MissingValues1}

#The total number of missing values in steps can be calculated using is.na() method to check whether the value is mising or not and then summing the logical vector.


missing_vals <- sum(is.na(rdata$steps))



#The missing

missing_vals

```

There are 2304 missing values in the dataset.


### 4.2 Strategy for filling in all of the missing values in the dataset

```{r MissingValues2}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}


```

###4.3 Fill Missing Values

```{r FillMissing}
rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)

str(rdata_fill)

head(rdata_fill)


sum(is.na(rdata_fill$steps))

```

###4.4 A histogram of the total number of steps taken each day

Now let us plot a histogram of the daily total number of steps taken, plotted with a bin interval of 1000 steps, after filling missing values

```{r HistogramPatterns}

fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram

ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 

```


# 4.5 Calculate and report the mean and median total number of steps taken per day.

```{r NewMeanMedian}


steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)

steps_mean_fill

steps_median_fill

```

Mean number of steps taken per day is 1.076618910^{4}.

Median number of steps taken per day is 1.076618910^{4}.

Since I imputed the missing values by the mean number of steps per interval, there is no difference in mean before and after imputing that is not surprising. The median has changed a little bit.



#5. Are there differences in activity patterns between weekdays and weekends?

## 5.1 Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r DiffBtwnPatterns}

weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)

```

### 5.2 Panel Plot

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r Plot1}

ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()

```

